Labor Market Income and Substitution Effects Explain Contemporary Natural Selection

### 2020-12-17 why-negative-selection.Rmd
Improve intro and conclusion. Mention Fermi paradox.

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Geneticists
have long hypothesized that natural selection is taking place in modern human
populations. Recent work confirms this using genome-wide analysis [@Barban_2016;
@beauchamp2016genetic;
@kong2017selection; @sanjak2018evidence]. In particular, genetic variants
associated with higher educational attainment are being selected against,
although effect sizes appear small.

@@

Because selection effects are concentrated in poorer groups, they may also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3), for children in households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with 
educational attainment, which lowers the mean, is stronger at the bottom of the 
income distribution, and this increases the differences between groups.

@@

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in 
modern societies.
Economics possesses a range of theoretical models on the effects of skills,
education and income [see @hotz1997economics; @lundberg2007american]. One
perennial problem is how to test these theories in a world where education,
labour and marriage markets all interact. Genetic data, such as polygenic
scores, could help to pin down directions of causality. Conversely, theory 
and empirical results from economics can shine a light on the mechanisms 
behind natural selection.

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
polygenic scores between the highest and lowest income groups. Column "Actual"
shows respondents' scores weighted by Age/Qualification times number of
children. Column "Without selection" shows scores weighted by Age/Qualification
only, i.e. if all couples had the same number of children.

>>>>

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how
individual differences in complex traits and disease risk arise. Geneticists
have long hypothesized that natural selection is taking place in modern human
populations [@galton1869hereditary;@darwin1871the]; recent work confirms this 
using genome-wide analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

Because selection effects are concentrated in poorer groups, they may also increase
inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3), for children in households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with 
educational attainment, which lowers the mean, is stronger at the bottom of the 
income distribution, and this increases the differences between groups.  Overall,
natural selection increases inequality for 29 out of 33 polygenic scores 
(Appendix Table \@ref(tab:tbl-inequality-effects)).

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism", including speculation that the Fermi paradox may be resolved by
dysgenic selection among alien species [@sarraf2019modernity]. Others are more
sanguine, and argue that natural selection's effects are outweighed by
environmental improvements, such as those underlying (e.g.) the Flynn effect
[@flynn1987massive]. The evidence here may add some nuance to this debate.
Patterns of natural selection have been relatively consistent across the past two
generations, but they are not the outcome of a universal, society-wide phenomenon.
Instead they result from a sum of opposing forces, operating in different
parts of society and pulling in different directions.

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in
modern societies. Economics possesses a range of theoretical models on the
effects of skills, education and income [see @hotz1997economics;
@lundberg2007american]. One perennial problem is how to test these theories in a
world where education, labour and marriage markets all interact. Genetic data,
such as polygenic scores, could help to pin down the direction of causality.
Conversely, theory and empirical results from economics can shine a light on the
mechanisms behind natural selection, and thereby on the nature of individual
differences in complex traits and disease risk.

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
polygenic scores between the highest and lowest income groups. Column "Actual"
uses respondents' scores weighted by Age/Qualification times number of
children. Column "Without selection" uses scores weighted by Age/Qualification
only, i.e. if all couples had the same number of children.
